Ages of Stars and Planets in the Kepler Field Over the First Four Billion Years

### 2024-08-19 papers/paper/ms.tex
update ms

% upon AAS submission
\documentclass[11pt,twocolumn,tighten,linenumbers]{aastex63}
%\documentclass[12pt,twocolumn,tighten,linenumbers,trackchanges]{aastex63}
% drafting / arxiv
%\documentclass[11pt,twocolumn,tighten]{aastex63}
\turnoffedit

@@

\begin{figure*}[!t]
	\begin{center}
		\subfloat{
			\includegraphics[width=0.47\textwidth]{prot_teff_Santos19_Santos21_dquality.pdf}
		}
		\subfloat{
			\includegraphics[width=0.47\textwidth]{koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.pdf}
		}
	\end{center}
	\vspace{-0.5cm}
	\caption{
    {\bf Rotation periods for Kepler target stars (left) and known
    planet hosts (right)}.  The gray lines are ``mean fits'' to the
    rotation sequences of open clusters.  The 2-D histogram of the
    stellar sample (left) includes only apparently single stars near
    the main sequence with $\log g$$>$4.2, RUWE$<$1.4, and
    temperatures of 3800--6200\,K.  The planet-hosts (right) require
    the same stellar cuts, and include only the confirmed and
    candidate planets described in Section~\ref{subsec:plflags}.  The
    number of stars $N_{\rm s}$ and planets $N_{\rm p}$ are noted in
    the relevant panels.  A simple estimate for the number of young
    planets discovered by Kepler follows by counting points below the
    mean rotational isochrones.
	}
	\label{fig:prot_vs_teff}
\end{figure*}

@@

Finally, since our scope is focused on rotation-based ages, we
restricted our attention to stars with reported $P_{\rm
rot}$$<$45\,days.  The slowest-rotating FGK stars in the open clusters
used to calibrate our gyrochronology model have $P_{\rm
rot}$$\approx$35\,days.  Figure~\ref{fig:prot_vs_teff} shows a subset
of the resulting \nuniqstarsantosrot\ Kepler stars with rotation
periods compiled from \citetalias{Santos_2019},
\citetalias{Santos_2021}, and our extended KOI list.

To assess the statistical uncertainties of these rotation periods, we
compared our adopted periods with those reported by
\citet{McQuillan_2014}.  The details are in
Appendix~\ref{app:reinhold23}.  We found that for $P_{\rm
rot}\lesssim$15\,days, the two datasets agree at a precision of
$\lesssim$0.01$P_{\rm rot}$.  At longer periods of $P_{\rm
rot}\approx$30\,days, the agreement was typically at the
$\lesssim$0.03$P_{\rm rot}$ level, and the envelope of the period
difference increased roughly linearly with period.  Based on this
comparison, we adopted a simple prescription for the period
uncertainties, such that there are 1\% relative uncertainties below
$P_{\rm rot}=15$\,days, and a linear increase thereafter, with slope
set to require 3\% $P_{\rm rot}$ uncertainties at rotation periods of
30 days.

@@

We calculated gyrochrone ages using \texttt{gyro-interp}
\citep{Bouma_2023}, which is a method designed to address the fact
that stars with the same mass and same rotation period can have a wide
range of ages \citep[e.g.][]{Curtis_2019_ngc6811}.  The gyrochrone age
posterior should therefore incorporate the intrinsic population-level
scatter into its statistical precision.  Figure~\ref{fig:models}
highlights this problem, particularly in regions where the 0.1--1\,Gyr
stars overlap.

@@

{\it Near the main sequence (bit 9)}---Figure~\ref{fig:stellarprops} shows
that many stars observed by Kepler
are far from the main sequence.  Some of the challenges this introduces
for rotation-based ages include unresolved binaries, metallicity,
reddening, and drivers of rotation other than magnetic braking.
After exploring various options, we settled on the orange locus in the
$\log g$--$T_{\rm eff}$ plane shown in Figure~\ref{fig:stellarprops}
as a way of flagging unresolved binaries, as well as evolved
late-F and early-G stars.  While the exact details of how this locus
is constructed are arbitrary (see Appendix~\ref{app:linemethod}), the
general aim is to flag stars for which there is evidence based on
their location in the HR diagram that our gyrochronology model may not
be reliable.  While the $\log g$--$T_{\rm eff}$ cut flags about half
of Kepler stars with rotation periods as potentially not being
suitable for gyrochronology, a few outliers in $M_{\rm G}$ vs.~$G_{\rm
BP}$-$G_{\rm RP}$ remained after applying it.  Such outliers are
likely also questionable.  We therefore also fitted a polynomial to
the KOI main sequence in $M_{\rm G}$ vs.~$G_{\rm
	BP}$-$G_{\rm RP}$, and flagged stars more than 1\,magnitude from
this locus as part of the same bitflag.

@@

We were interested in not only the overall age distribution of the
Kepler field, but also in the reliability of individual ages for young
stars known to host planets.  We attempted to assess this reliability using
quality bitmasks (Table~\ref{tab:planets}), a comparison against open
cluster ages (Section~\ref{sec:litagecomparison}), and a comparison
between rotation and lithium ages.  The open cluster comparison
suggested that our ages were generally accurate at $\ll$0.5\,dex
across 0.04--2.5\,Gyr, comparable to their quoted precision.
Nonetheless, the upper-left panel of Figure~\ref{fig:agescalecompone}
does show outliers, typically $\lesssim$10\% of the population
in any cluster.  These stars could be either field star contaminants,
or else anomalous stars whose physical rotation histories were altered
by processes not captured by our statistical uncertainties.
Similarly, although we found $t_{\rm gyro}$ and $t_{\rm Li}$ to agree
for \ltonegyrhighqconfirmedtwosided\ planets orbiting ``high-quality''
stars, we did find one system, Kepler-786, with radically different
lithium (\kepseveneightsix\,Myr) and rotational
(\kepseveneightsixgyro\,Myr) ages.  Planetary mergers
are one process that could produce such a signal, but
testing this would require a method for measuring differential abundances
across a large number of elements.

@@

We argued in Section~\ref{subsec:rotsel} that adopting
the periods from \citet{Santos_2019,Santos_2021} yielded the best
possible balance between homogeneity and sensitivity for both KOIs and
for the broader Kepler sample.  Nonetheless, it is interesting to
consider what would happen if we were to analyze only the rotation
periods from \citet{McQuillan_2014}.
Figures~\ref{fig:prot_vs_teff_mcq} and~\ref{fig:hist_tgyro_mcq} repeat
the analysis, but make this alternate choice.
Figure~\ref{fig:prot_vs_teff_mcq} shows that while the overall
contours of the $P_{\rm rot}$ vs.~$T_{\rm eff}$ distributions are
similar, the \citeauthor{Santos_2019} distribution includes more
stars, particularly at longer rotation periods.  This is connected to
a stronger noise-dependent cutoff in the \citet{McQuillan_2014} sample
than in the \citeauthor{Santos_2019} samples (see
\citet{Masuda2022_amplevoln} Figures~10 and 15).
Figure~\ref{fig:hist_tgyro_mcq} shows the impact on the inferred
rotation-based age histograms: while the \citeauthor{Santos_2019}
distribution peaks at 2.8-3.0\,Gyr, the \citeauthor{McQuillan_2014}
distribution peaks at 2.4-2.6\,Gyr.  A number of metrics shift.  For
instance, the ratio of the star formation rate 3\,Gyr age to today is
\ratiosfr$\pm$\uncratiosfr\ assuming the \citeauthor{Santos_2019}
periods, and \mcquillanonlyratiosfr$\pm$\mcquillanonlyuncratiosfr\
assuming the \citeauthor{McQuillan_2014} periods.  Similarly, the
ratio of ``old'' (2-3\,Gyr) to ``young'' (0-1\,Gyr) stars shifts from
\ratioobtoybstars\ in the \citeauthor{Santos_2019} case to
\mcquillanonlyratioobtoybstars\ in the \citeauthor{McQuillan_2014}
case.  The K dwarf age distributions (Figure~\ref{fig:hist_tgyro_mcq}
lower panel) however appear quite similar.  Generally, these plots
support the conclusion that the $t_{\rm gyro}$ age distribution in the
Kepler field is non-uniform, with a peak in the 2.5-3\,Gyr range.
Nonetheless, some of the details do shift depending on the choice of
rotation detection pipeline.

>>>>

% upon AAS submission
\documentclass[11pt,twocolumn,tighten,linenumbers]{aastex63}
%\documentclass[11pt,twocolumn,tighten,linenumbers,trackchanges]{aastex63}
% drafting / arxiv
%\documentclass[11pt,twocolumn,tighten]{aastex63}
\turnoffedit

@@

\begin{figure*}[!t]
	\begin{center}
		\subfloat{
			\includegraphics[width=0.47\textwidth]{prot_teff_Santos19_Santos21_dquality.pdf}
		}
		\subfloat{
			\includegraphics[width=0.47\textwidth]{koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.pdf}
		}
	\end{center}
	\vspace{-0.5cm}
	\caption{
    {\bf Rotation periods for Kepler target stars (left) and known
    planet hosts (right)}.  The gray lines are \replaced{``mean fits'' to the
    rotation sequences of open clusters}{seventh-order polynomials
    fitted to the slow rotation sequences of open clusters \citep[see Section~3.3 of][]{Bouma_2023}}.  The 2-D histogram of the
    stellar sample (left) includes only apparently single stars near
    the main sequence with $\log g$$>$4.2, RUWE$<$1.4, and
    temperatures of 3800--6200\,K.  The planet-hosts (right) require
    the same stellar cuts, and include only the confirmed and
    candidate planets described in Section~\ref{subsec:plflags}.  The
    number of stars $N_{\rm s}$ and planets $N_{\rm p}$ are noted in
    the relevant panels.  A simple estimate for the number of young
    planets discovered by Kepler follows by counting points below the
    \deleted{mean }rotational isochrones.
	}
	\label{fig:prot_vs_teff}
\end{figure*}

@@

Finally, since our scope is focused on rotation-based ages, we
restricted our attention to stars with reported $P_{\rm
rot}$$<$45\,days.  The slowest-rotating FGK stars in the open clusters
used to calibrate our gyrochronology model have $P_{\rm
rot}$$\approx$35\,days.  Figure~\ref{fig:prot_vs_teff} shows a subset
of the resulting \nuniqstarsantosrot\ Kepler stars with rotation
periods compiled from \citetalias{Santos_2019},
\citetalias{Santos_2021}, and our extended KOI list.\added{  The gray
lines in this figure are the polynomials described by \citet{Bouma_2023}
that were
fitted to the slow rotation sequences of the Pleiades \citep{Rebull_2016},
Praesepe \citep{Rampalli_2021}, NGC-6811 \citep{Curtis_2019_ngc6811}, NGC-6819 \citep{Meibom_2015},
Ruprecht-147 \citep{Curtis_2020}, and M67 \citep{Barnes2016,Dungee_2022,Gruner_2023}.}

To assess the statistical uncertainties of \replaced{these}{our adopted} rotation periods, we
compared our \deleted{adopted} periods with those reported by
\citet{McQuillan_2014}.  The details are in
Appendix~\ref{app:reinhold23}.  We found that for $P_{\rm
rot}\lesssim$15\,days, the two datasets agree at a precision of
$\lesssim$0.01$P_{\rm rot}$.  At longer periods of $P_{\rm
rot}\approx$30\,days, the agreement was typically at the
$\lesssim$0.03$P_{\rm rot}$ level, and the envelope of the period
difference increased roughly linearly with period.  Based on this
comparison, we adopted a simple prescription for the period
uncertainties, such that there are 1\% relative uncertainties below
$P_{\rm rot}=15$\,days, and a linear increase thereafter, with slope
set to require 3\% $P_{\rm rot}$ uncertainties at rotation periods of
30 days.

@@

We calculated gyrochrone ages using \texttt{gyro-interp}
\citep{Bouma_2023}, which is a method designed to address the fact
that stars with the same mass and same rotation period can have a wide
range of ages \citep[e.g.][]{Gallet_Bouvier_2015}.  The gyrochrone age
posterior should therefore incorporate the intrinsic population-level
scatter into its statistical precision.  Figure~\ref{fig:models}
highlights this problem, particularly in regions where the 0.1--1\,Gyr
stars overlap.

@@

{\it Near the main sequence (bit 9)}---Figure~\ref{fig:stellarprops} shows
that many stars observed by Kepler
are far from the main sequence.  Some of the challenges this introduces
for rotation-based ages include unresolved binaries, metallicity,
reddening, and drivers of rotation other than magnetic braking.
After exploring various options, we settled on the orange locus in the
$\log g$--$T_{\rm eff}$ plane shown in Figure~\ref{fig:stellarprops}
as a way of flagging unresolved binaries, as well as evolved
late-F and early-G stars.  While the exact details of how this locus
is constructed are arbitrary (see Appendix~\ref{app:linemethod}), the
general aim is to flag stars for which there is evidence based on
their location in the \replaced{HR}{Kiel} diagram that our gyrochronology model may not
be reliable.  While the $\log g$--$T_{\rm eff}$ cut flags about half
of Kepler stars with rotation periods as potentially not being
suitable for gyrochronology, a few outliers in $M_{\rm G}$ vs.~$G_{\rm
BP}$-$G_{\rm RP}$ remained after applying it.  Such outliers are
likely also questionable.  We therefore also fitted a polynomial to
the KOI main sequence in $M_{\rm G}$ vs.~$G_{\rm
	BP}$-$G_{\rm RP}$, and flagged stars more than 1\,magnitude from
this locus as part of the same bitflag.

@@

We were interested in not only the overall age distribution of the
Kepler field, but also in the reliability of individual ages for young
stars known to host planets.  We attempted to assess this reliability using
quality bitmasks (Table~\ref{tab:planets}), a comparison against \deleted{open }
cluster ages (Section~\ref{sec:litagecomparison}), and a comparison
between rotation and lithium ages.  The \deleted{open }cluster comparison
suggested that our ages were generally accurate at $\ll$0.5\,dex
across 0.04--2.5\,Gyr, comparable to their quoted precision.
Nonetheless, the upper-left panel of Figure~\ref{fig:agescalecompone}
does show outliers, typically $\lesssim$10\% of the population
in any cluster.  These stars could be either field star contaminants,
or else anomalous stars whose physical rotation histories were altered
by processes not captured by our statistical uncertainties.
Similarly, although we found $t_{\rm gyro}$ and $t_{\rm Li}$ to agree
for \ltonegyrhighqconfirmedtwosided\ planets orbiting ``high-quality''
stars, we did find one system, Kepler-786, with radically different
lithium (\kepseveneightsix\,Myr) and rotational
(\kepseveneightsixgyro\,Myr) ages.  Planetary mergers
are one process that could produce such a signal, but
testing this would require a method for measuring differential abundances
across a large number of elements.

@@

We argued in Section~\ref{subsec:rotsel} that adopting
the periods from \citet{Santos_2019,Santos_2021} yielded the best
possible balance between homogeneity and sensitivity for both KOIs and
for the broader Kepler sample.  Nonetheless, it is interesting to
consider what would happen if we were to analyze only the rotation
periods from \citet{McQuillan_2014}.
Figures~\ref{fig:prot_vs_teff_mcq} and~\ref{fig:hist_tgyro_mcq} repeat
the analysis, but make this alternate choice.
Figure~\ref{fig:prot_vs_teff_mcq} shows that while the overall
contours of the $P_{\rm rot}$ vs.~$T_{\rm eff}$ distributions are
similar, the \citeauthor{Santos_2019} distribution includes more
stars, particularly at longer rotation periods.  This is connected to
a stronger noise-dependent cutoff in the \citet{McQuillan_2014} sample
than in the \citeauthor{Santos_2019} samples (see
\citealt{Masuda2022_amplevoln} Figures~10 and 15).
Figure~\ref{fig:hist_tgyro_mcq} shows the impact on the inferred
rotation-based age histograms: while the \citeauthor{Santos_2019}
distribution peaks at 2.8-3.0\,Gyr, the \citeauthor{McQuillan_2014}
distribution peaks at 2.4-2.6\,Gyr.  A number of metrics shift.  For
instance, the ratio of the star formation rate 3\,Gyr age to today is
\ratiosfr$\pm$\uncratiosfr\ assuming the \citeauthor{Santos_2019}
periods, and \mcquillanonlyratiosfr$\pm$\mcquillanonlyuncratiosfr\
assuming the \citeauthor{McQuillan_2014} periods.  Similarly, the
ratio of ``old'' (2-3\,Gyr) to ``young'' (0-1\,Gyr) stars shifts from
\ratioobtoybstars\ in the \citeauthor{Santos_2019} case to
\mcquillanonlyratioobtoybstars\ in the \citeauthor{McQuillan_2014}
case.  The K dwarf age distributions (Figure~\ref{fig:hist_tgyro_mcq}
lower panel) however appear quite similar.  Generally, these plots
support the conclusion that the $t_{\rm gyro}$ age distribution in the
Kepler field is non-uniform, with a peak in the \replaced{2.5}{2.3}-3\,Gyr range.
Nonetheless, some of the details do shift depending on the choice of
rotation detection pipeline.
